Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-06-17 ICML/ARTEMISS_proposal_Oberman_vanBuuren_Vink.tex

\begin{abstract}
Iterative imputation is a popular tool to accommodate missing data. While it is widely accepted that valid inferences can be obtained with this technique, these inferences all rely on algorithmic convergence. There is no consensus on how to evaluate the convergence properties of the method. This paper provides insight into identifying non-convergence in iterative imputation algorithms. Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by diagnostic methods. We conclude that it never hurts to iterate longer, but such calculations hardly bring added value.
\end{abstract}

@@

We induce non-convergence in the imputation algorithm with two sets of simulation conditions: early stopping and missingness severity. Early stopping implies that we vary the number of iterations before terminating the imputation algorithm (between 1 and 100 iterations). The missingness severity is determined\footnote{I.e., we only consider a `missing completely at random' missingness mechanism \citep{rubin76}.} by the proportion of incomplete cases (between 5 and 95\%). The simulation set-up is summarized in pseudo-code (see Algorithm \ref{alg:setup}), while the complete script and technical details are available from \href{https://github.com/hanneoberman/MissingThePoint}{github.com/hanneoberman/MissingThePoint}.

>>>>

\begin{abstract}
Iterative imputation is a popular tool to accommodate missing data. While it is widely accepted that valid inferences can be obtained with this technique, these inferences all rely on algorithmic convergence. There is no consensus on how to evaluate the convergence properties of the method. Our study provides insight into identifying non-convergence in iterative imputation algorithms. We found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by diagnostic methods. We conclude that it never hurts to iterate longer, but such calculations hardly bring added value.
\end{abstract}

@@

To induce non-convergence in the imputation algorithm we use  two sets of simulation conditions: early stopping and missingness severity. Early stopping implies that we vary the number of iterations before terminating the imputation algorithm (between 1 and 100 iterations). The missingness severity is determined\footnote{I.e., we only consider a `missing completely at random' missingness mechanism \citep{rubin76}.} by the proportion of incomplete cases (between 5 and 95\%). We provide a summary of the simulation set-up in Algorithm \ref{alg:setup}, whereas the complete script and technical details are available from \href{https://github.com/hanneoberman/MissingThePoint}{github.com/hanneoberman/MissingThePoint}.
